Missing The Point

### 2020-10-19 Manuscript.Rmd
Clarify results and grammar check

---
title: "Missing The Point"
author: "Hanne Oberman"
date: "7-9-2020"
output: html_document
bibliography: ref.bib
---

@@

# sim results plotting function
generic_plot <- function(v, name, ...){
  results %>%
    mutate(
      p = factor(p, levels = c(0.25, 0.5, 0.75), labels = c("25%", "50%", "75%")),
      mech = factor(mech, levels = c("MCAR", "MAR", "MNAR")),
      est = est - 1,
      rsq = rsq - 0.5
      ) %>% 
  ggplot(aes(
    x=.data[["it"]], 
    y=.data[[v]], 
    color = .data[["p"]], 
    shape = .data[["mech"]], 
    linetype = .data[["mech"]]
    )) +
  geom_line(na.rm = TRUE) +
  labs(
    x = "Number of iterations",
    y = name,
    color = "Incomplete \n cases", 
    shape = "Missingness \n mechanism", 
    linetype = "Missingness \n mechanism") +
  theme_classic() + 
  scale_color_manual(values = c(
    #"#009E73", #green
    "#56B4E9", #blue
    #"#E69F00", #orange
    "#F0E442", #yellow
    "#D55E00"  #red
  ))
  # scale_color_manual(values = c(
  #   # '#228833', #green
  #   # '#66CCEE', #blue
  #   '#CCBB44', #yellow
  #   '#EE6677',  #red~ish
  #   '#AA3377'  #maroon
  #   ))
}
```

@@

@@

Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms, since there is no consensus on how to evaluate the convergence properties currently. We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

In this paper we explore diagnostic methods for iterative imputation algorithms. For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p. 147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p. 5). Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.

We propose a novel multivariate parameter to check for non-convergence in iterative algorithms. Our aim is to show which method is the most informative about non-convergence in iterative imputation. [explain that method = parameter + diagnostic + interpretation]

@@

-   What does non-convergence look like? --\> non-stat + non-mix in a certain parameter, e.g., chain means.

-   What are the consequences? --\> bias, under-coverage --\> Refer to van Buuren (2018) instead of reproducing this

-   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N\_eff is estimated using varcov matrix of the set of samples, while this is not the case in it im per se.

-   We consider autocorr and rhat, and monitor four parameters: chain means, chain variances, the scientific estimand, and lambda.

## Simulation

@@

Algorithm 1: simulation set-up (pseudo-code)

@@

Our aim is to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice]. Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

Data are generated in each simulation repetition for a sample of $n_{obs} = 1000$ cases, before inducing missingness. Let the complete set be three multivariately normal random variables

@@

The complete set is amputed according to the nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases. [Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences). And add the `ampute` function?]

@@

We use a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.

@@

As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($\bar{Q}_{LL} - \bar{Q}_{UL}$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands.

## Results

We present the results of our simulation study visually because of the many simulation conditions. Within each plot, the results are split according to the missingness conditions (missingness mechanisms as line types, and proportion of incomplete cases as colors).

### Non-Convergence Identifiers

The next four plots show the autocorrelations in each parameter plotted against the number of iterations.

```{r echo=FALSE}
p <- generic_plot(v="ac.max.mu.Y", name = "AC chain means")
p %>% plotly::ggplotly()
```

This plot shows that irrespective of the missingness condition, autocorrelation in the chain means decreases rapidly until $it \geq 6$. This means that there is some initial trending within chains, but the stationarity does not improve substantively after the first few iterations.

```{r echo=FALSE}
p <- generic_plot(v="ac.max.sigma.Y", name = "AC chain variances")
p %>% plotly::ggplotly() 
```

Autocorrelations in the chain variances do not show a systematic effect of the missingness conditions either. The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous, but roughly when $it \geq 10$.

```{r echo=FALSE}
p <- generic_plot(v="ac.max.qhat", name = "AC scientific estimate")
p %>% plotly::ggplotly() 
```

We see that the autocorrelations in the scientific estimates are more pronounced than in the univariate parameters (chain means and chain variances), especially when the proportion of missing cases is moderate to extreme. We observe the worst stationarity in conditions with 75% of cases having missing values. There is no clear effect of the missingness mechanisms. Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$.

```{r echo=FALSE}
p <- generic_plot(v="ac.max.lambda", name = "AC novel theta")
p %>% plotly::ggplotly() 
```

Autocorrelations in the novel parameter show a similar trend to the autocorrelations in the scientific estimates. Around $it \geq 20$ is required before trending in this parameter diminishes.

The next four plots show the $\widehat{R}$ in the different parameters.

```{r echo=FALSE}
p <- generic_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
p %>% plotly::ggplotly() 
```

In this plot we observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations---with the exception of an initial increase when $3 < it < 5$. There is no apparent differentiation between the missingness conditions. The mixing between chain means generally improves until $it \geq 30$ to $40$.

```{r echo=FALSE}
p <- generic_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
p %>% plotly::ggplotly() 
```

@@

```{r echo=FALSE}
p <- generic_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
p %>% plotly::ggplotly() 
```

With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again. We do, however, see some differentiation between missingness conditions. Conditions with 75% of cases being incomplete show more extreme non-mixing. The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.

```{r echo=FALSE}
p <- generic_plot(v="r.hat.max.lambda", name = "R hat novel theta")
p %>% plotly::ggplotly() 
```

@@

[These rhat plots all show some initialization before the fifth iteration: is rhat usefull before that??]

@@

```{r echo=FALSE}
p <- generic_plot(v = "est", name = "Bias in regression estimate") + 
  list(geom_hline(yintercept = 0, color = "grey")) 
p %>% plotly::ggplotly()
```

Bias in estimate depends mostly on the amount of missingness (less info to estimate the relation??). Overall 75% incomplete cases performs the worst, but is stable for $t \geq 6$.

@@

```{r echo=FALSE}
p <- generic_plot( v ="cov", name = "Coverage rate") + 
  list(geom_hline(yintercept = 0.95, color = "grey"))
p %>% plotly::ggplotly()

# p <- generic_plot( v ="cov", name = "Coverage rate") + 
#   list(geom_hline(yintercept = 0.95, color = "grey")) +
#   ylim(0.85, 1)

@@

Cov is more or less fine, except for MNAR with only 0.05% incomplete cases --\> due to bias in the estimate in combination with very narrow CI (see CIW!). However, all stable after it=3??

```{r echo=FALSE}
p <- generic_plot(v="CIW", name = "CI width") 
p %>% plotly::ggplotly()
```

CIW clearly determined by the proportion of incomplete cases. Worst for p=0.95, then stable after it 7.

```{r echo=FALSE}
p <- generic_plot(v="rsq", name = "Bias in coefficient of determination") +
  list(geom_hline(yintercept = 0, color = "grey"))
p %>% plotly::ggplotly()
```

Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=5.

@@

- need for multivar parameters!!

-   Estimates etc do not improve after it = 7

-   Convergence diagnostics keep improving, but are all stable after 20-30 it

-   Univariate thetas do not differentiate between missingness mechanisms or proportions of incomplete cases

@@

methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate conv slower than the est

calc rhat and cov for each block of 5 it

disc: conf val is likely to happen much sooner

as soon as we start down-weighting the first few it from the calc, the memory effect would

>>>>

---
title: "Missing The Point"
author: "Hanne Oberman"
date: "7-9-2020"
output:
  word_document: default
  pdf_document: default
  html_document: default
bibliography: ref.bib
---

@@

# sim results plotting function
simulation_plot <- function(v, name, ...){
  results %>%
    mutate(
      p = factor(p, levels = c(0.25, 0.5, 0.75), labels = c("25%", "50%", "75%")),
      mech = factor(mech, levels = c("MCAR", "MAR", "MNAR")),
      est = est - 1,
      rsq = rsq - 0.5
      ) %>% 
  ggplot(aes(
    x=.data[["it"]], 
    y=.data[[v]], 
    color = .data[["p"]], 
    shape = .data[["mech"]], 
    linetype = .data[["mech"]]
    )) +
  geom_line(na.rm = TRUE) +
  labs(
    x = "Number of iterations",
    y = name,
    color = "Incomplete \n cases", 
    shape = "Missingness \n mechanism", 
    linetype = "Missingness \n mechanism") +
  theme_classic() + 
  scale_color_manual(values = c(
    #"#009E73", #green
    "#56B4E9", #blue
    #"#E69F00", #orange
    "#F0E442", #yellow
    "#D55E00"  #red
  ))
  # scale_color_manual(values = c(
  #   # '#228833', #green
  #   # '#66CCEE', #blue
  #   '#CCBB44', #yellow
  #   '#EE6677',  #red~ish
  #   '#AA3377'  #maroon
  #   ))
}
```

@@

*Notation in this manuscript: *

- square brackets are comments for myself/stuff I need to review

- bullet points are things I need to expand

@@

Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms since there is no consensus on how to evaluate the convergence properties currently. We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

In this paper, we explore diagnostic methods for iterative imputation algorithms. For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p. 147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p. 5). Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.

We propose a novel multivariate parameter to check for non-convergence in iterative algorithms. We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation].

@@

-   What does non-convergence look like? --\> non-stationarity + non-mixing in a certain parameter, e.g., chain means.

-   What are the consequences? --\> bias, under-coverage --\> refer to van Buuren (2018) instead of reproducing this

<!-- -   Existing non-convergence identifiers for MCMC algorithms are not all applicable, e.g. N\_eff is estimated using the variance-covariance matrix of the set of samples, while this is not the case in iterative imputation per se. -->

-   We consider autocorrelation and rhat, and monitor four parameters: chain means, chain variances, a scientific estimate, and lambda.

- Explain lambda

## Simulation Set-Up

@@

**Algorithm 1: simulation set-up (pseudo-code)**

@@

We aim to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice]. Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

Data are generated in each simulation repetition for a complete set of $n_{obs} = 1000$ cases (i.e., before inducing missingness). We define three multivariately normal random variables, let

@@

[Add mvtnorm package here??] The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases. [Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences). And add the `ampute` function?]

@@

We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.

@@

As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($E(\bar{Q}_{LL} - \bar{Q}_{UL})$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!! and check if definition of ciw is correct].

## Simulation Results

The following plots provide an overview of the results for each method and each performance measure, contrasted to the number of iteration in the imputation algorithm. Within the plots, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].

### Diagnostic Methods

<!-- The next four plots show the autocorrelations in each parameter plotted against the number of iterations.  -->

```{r echo=FALSE}
p <- simulation_plot(v="ac.max.mu.Y", name = "Autocorrelation chain means")
p # %>% plotly::ggplotly()
```

In this plot, our diagnostic method is the amount of autocorrelation in the chain means. We can interpret this as the extent to which the average imputed value trends across iterations. We see that autocorrelation in the chain means rapidly decreases until $it \geq 6$. This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations. These results hold irrespective of the missingness condition.

```{r echo=FALSE}
p <- simulation_plot(v="ac.max.sigma.Y", name = "Autocorrelation chain variances")
p # %>% plotly::ggplotly() 
```

Autocorrelation in the chain variances shows something similar. The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally when $it \geq 10$. We do not observe a systematic effect of the missingness conditions here either.

```{r echo=FALSE}
p <- simulation_plot(v="ac.max.qhat", name = "Autocorrelation scientific estimate")
p # %>% plotly::ggplotly() 
```

There is more autocorrelation in the scientific estimates than in the univariate parameters (chain means and chain variances). We observe the highest autocorrelations in conditions where 75% of cases are incomplete. Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$. There is no clear effect of the missingness mechanisms.

```{r echo=FALSE}
p <- simulation_plot(v="ac.max.lambda", name = "Autocorrelation novel theta")
p # %>% plotly::ggplotly() 
```

The autocorrelation in the novel parameter exhibits a similar trend to the autocorrelation in the scientific estimates. Trending in this parameter diminishes when $it \geq 20$.

<!-- The next four plots show the $\widehat{R}$ in the different parameters. -->

```{r echo=FALSE}
p <- simulation_plot(v="r.hat.max.mu.Y", name = "R hat chain means")
p # %>% plotly::ggplotly() 
```

We observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations. An exception to this observation is the initial increase when $3 < it < 5$. According to these results, the mixing between chain means generally improves until $it \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

```{r echo=FALSE}
p <- simulation_plot(v="r.hat.max.sigma.Y", name = "R hat chain variances")
p # %>% plotly::ggplotly() 
```

@@

```{r echo=FALSE}
p <- simulation_plot(v="r.hat.max.qhat", name = "R hat scientific estimate")
p # %>% plotly::ggplotly() 
```

With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again. We do, however, see some differentiation between missingness conditions. Conditions where 75% of the cases are incomplete show more extreme non-mixing. The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.

```{r echo=FALSE}
p <- simulation_plot(v="r.hat.max.lambda", name = "R hat novel theta")
p # %>% plotly::ggplotly() 
```

@@

[These rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]

@@

```{r echo=FALSE}
p <- simulation_plot(v = "est", name = "Bias in regression estimate") + 
  list(geom_hline(yintercept = 0, color = "grey")) 
p # %>% plotly::ggplotly()
```

We see that within a few iterations the average bias approaches zero. When $it \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].

@@

```{r echo=FALSE}
p <- simulation_plot( v ="cov", name = "Coverage rate") + 
  list(geom_hline(yintercept = 0.95, color = "grey"))
p # %>% plotly::ggplotly()

# p <- simulation_plot( v ="cov", name = "Coverage rate") + 
#   list(geom_hline(yintercept = 0.95, color = "grey")) +
#   ylim(0.85, 1)

@@

Nominal coverage is quickly reached. After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].

```{r echo=FALSE}
p <- simulation_plot(v="CIW", name = "Average CI width") 
p # %>% plotly::ggplotly()
```

The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached. Depending on the proportion of incomplete cases this takes up-to $it \geq 9$.

```{r echo=FALSE}
p <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
  list(geom_hline(yintercept = 0, color = "grey"))
p # %>% plotly::ggplotly()
```
Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations. We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].

<!-- Magnitude of the bias in $r^2$ perfectly follows the missingness mechanisms: least bias for MCAR, most for MNAR. Worst effect of non-convergence for $p=0.95$, MNAR, but stable from it=6. -->

@@

-   Convergence diagnostics keep improving substantially until it = 20-30

-   Performance measures do not improve after it = 9

-  [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]

-   Univariate thetas may under-estimate non-convergence.

@@

- Idea: calculate rhat for each block of 5 it
